\documentclass[11pt,a4paper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{float}

\setstretch{1.1}
\setlength{\parskip}{0.6em}
\setlength{\parindent}{0pt}

\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection}{0.5em}{}

\title{\textbf{SEMTM0045 Portfolio}\\Theme 2, Topic 1: Evaluation under Uncertainty and Limited Data}
\author{Your Name / Student ID}
\date{}

\begin{document}

\maketitle

\section{Abstract}

This portfolio addresses the methodological challenge of evaluating classical machine learning models under limited and imbalanced data conditions. The project is motivated by the concern that common validation procedures may generate unstable, overly optimistic, or insufficiently transparent performance estimates when sample size is small and uncertainty is high. The report critically reviews the literature on cross-validation, uncertainty estimation, selection bias, and imbalance-sensitive metric choice, and identifies a gap in the integration of these issues into a coherent practical framework for model assessment. In response, the portfolio proposes a comparative study design that evaluates common validation protocols with explicit attention to estimate stability, uncertainty-aware reporting, and metric appropriateness under class imbalance. The intended contribution is not a new predictive algorithm, but a more reliable and methodologically defensible framework for evaluating classical machine learning models in constrained-data settings.

\section{Introduction}

Reliable performance evaluation is central to machine learning research and practice. Reported model quality is often used to justify model selection, support claims about generalisability, and inform practical deployment. However, the credibility of such claims depends not only on the predictive model itself, but also on the evaluation procedure used to assess it. This issue becomes especially important when models are developed under constrained-data conditions, where limited sample size and class imbalance can make performance estimates unstable, overly optimistic, or difficult to interpret.

In many applied studies, model evaluation is still treated as a relatively routine stage of the workflow. A dataset is divided into training and testing subsets, a model is fitted, and one or more summary metrics are reported. While this process appears straightforward, the methodological literature suggests that such simplicity can be misleading. Cross-validation procedures differ substantially in their statistical properties, and their outputs should not be treated as interchangeable. Moreover, uncertainty in validation results is often much larger than is implied by a single reported score, especially when sample sizes are small or when resampling-based estimates are interpreted without sufficient caution.

The problem is compounded when class imbalance is present. In such settings, conventional metrics may fail to capture the practical quality of prediction, particularly for minority classes. A model may achieve apparently strong overall performance while still performing poorly on the cases that matter most. This means that evaluation design involves at least two interconnected questions: how data should be split or resampled, and how model performance should be measured and reported. Both questions become more difficult under limited-data conditions, where each observation has greater influence on the final estimate.

The literature also suggests that evaluation problems are not restricted to the final train--test comparison. Bias may enter earlier through the model selection process itself. If validation results are noisy, repeated tuning and comparison may lead to overfitting at the selection stage, even when the final model appears to have been evaluated appropriately. As a result, unreliable evaluation is not merely a statistical inconvenience; it is a methodological issue that can distort conclusions about which model should be trusted.

For these reasons, this portfolio focuses on the problem of \emph{evaluation under uncertainty and limited data}. Rather than proposing a new predictive algorithm, the report examines how common evaluation protocols behave under constrained-data conditions and whether more uncertainty-aware validation strategies can provide a more trustworthy basis for model comparison. The emphasis is therefore methodological rather than algorithmic. The project is concerned with how machine learning models should be judged when evidence is limited, variability is substantial, and apparently strong results may not be as reliable as they first appear.

\section{Critical Literature Review}

\subsection{Cross-validation as an evaluation framework}

A useful starting point for this topic is the recognition that cross-validation is not a single method, but a broad family of procedures with different goals and statistical properties. Arlot and Celisse (2010) provide one of the most comprehensive overviews of this landscape, showing that validation procedures differ in bias, variance, computational cost, and appropriateness for model selection or risk estimation. Their analysis is important because it challenges the common practical assumption that any cross-validation procedure is an acceptable generic solution. Instead, evaluation design must be understood as a methodological choice that depends on the structure of the problem, the available sample size, and the intended interpretation of the reported score.

This perspective is reinforced by more practice-oriented work such as Yates et al. (2022/2023), who argue that evaluation should not be reduced to selecting the highest mean validation score. Their discussion highlights issues such as estimation uncertainty, bias correction, and cautious decision rules in model comparison. Taken together, these studies establish an important baseline for the present portfolio: evaluation protocols should be selected deliberately rather than used by convention. This provides the foundation for a more critical discussion of why limited-data settings make this choice especially consequential.

\subsection{Uncertainty and instability in limited-data evaluation}

One of the strongest themes in the literature is that performance estimation becomes substantially less reliable when sample size is small. This is not simply because noise increases in an intuitive sense, but because the statistical behaviour of validation procedures becomes more fragile. Grandvalet and Bengio (2004) make this point at a fundamental level by proving that there is no universally unbiased estimator of the variance of k-fold cross-validation. This result is highly significant for the current topic because it demonstrates that uncertainty in model evaluation is not merely underreported in practice; it is also intrinsically difficult to estimate in a principled way. In other words, the problem is structural rather than cosmetic.

Nadeau and Bengio (2003) reach a closely related conclusion in their work on inference for the generalisation error. They show that naive treatment of repeated validation estimates as independent observations can underestimate variance and lead to unreliable statistical comparisons between algorithms. This matters because it suggests that even when researchers attempt to move beyond a single train--test split, the interpretation of repeated resampling results remains methodologically delicate. More data points in the output do not automatically translate into more trustworthy evidence.

These concerns become even more visible in small-sample applications. Varoquaux (2018) argues that cross-validation in small datasets can produce error bars large enough to undermine apparently strong conclusions. The key insight here is not simply that variability exists, but that the practical significance of that variability is often underappreciated. A model may appear to outperform another under one split or one validation protocol, yet the underlying uncertainty may be so large that the observed difference is not robust. This is particularly relevant for the present portfolio because it directly motivates a shift away from point-estimate thinking and towards uncertainty-aware evaluation.

Zantvoort et al. (2024) further extend this concern by showing that very small datasets can systematically inflate confidence in predictive performance. Their work suggests that limited data does not merely make estimates noisier; it can also create a systematic tendency towards over-optimistic interpretation, particularly when more complex models are used. This strengthens the case for a methodological framework that is not only variance-aware but also cautious about how results are communicated.

\subsection{Selection bias and over-optimistic evaluation}

A second major theme in the literature is that unreliable evaluation may arise not only from the final validation protocol, but also from the model selection process itself. Cawley and Talbot (2010) argue that model selection criteria can overfit noisy validation outcomes, creating selection bias that inflates reported performance. This contribution is especially important because it broadens the problem definition. Evaluation instability is not solely a matter of how the data is split at the end of the pipeline; it is also a consequence of how repeated comparisons and tuning decisions are made during development.

This argument has strong implications for studies conducted under limited-data conditions. When data are scarce, each validation estimate is more sensitive to sample composition, which increases the risk that model choice is driven by random fluctuations rather than genuine differences in predictive ability. As a result, evaluation cannot be treated as a final technical step. It must instead be treated as a full methodological process in which protocol choice, metric choice, and model comparison strategy interact.

\subsection{Imbalanced data and the role of metric choice}

While much of the methodological literature focuses on uncertainty and validation design, another important aspect of the topic concerns how model performance is measured. This becomes particularly important when class imbalance is present, which is common in many constrained-data applications. Saito and Rehmsmeier (2015) show that precision--recall analysis can be more informative than ROC-based evaluation in imbalanced binary classification. Their argument is highly relevant because it demonstrates that evaluation reliability depends not only on how data are resampled, but also on whether the chosen metric reflects the practical prediction goal.

This point is especially important for Topic 1. If a validation protocol produces a stable estimate of an inappropriate metric, the resulting evaluation may still be misleading. In other words, methodological reliability cannot be reduced to resampling design alone. Zhao et al. (2025) reinforce this broader perspective by surveying the small-sample imbalance problem as a joint methodological challenge rather than two separate technical inconveniences. Their review suggests that limited sample size and imbalance often interact in ways that require integrated treatment. This is directly aligned with the present portfolio, which seeks to connect uncertainty-aware validation with imbalance-sensitive metric selection.

\subsection{Applied validation design and practical relevance}

Theoretical concerns about instability and bias are supported by a growing body of applied evidence. Singh et al. (2021), working in cardiovascular imaging, compare multiple validation regimens and show that single split-sample validation produces considerably less stable performance estimates than repeated cross-validation or bootstrap-based approaches. This finding is important because it demonstrates that methodological warnings in the statistical literature have direct practical consequences in real small-data scenarios.

Similarly, Allgaier and Pryss (2024) show that evaluation design must reflect realistic deployment conditions, especially when hidden group structure or temporal drift is present. Their work broadens the discussion from internal validation stability to external validity and practical relevance. This is useful for the present report because it suggests that trustworthy evaluation is not only a question of statistical precision, but also of whether the testing design meaningfully reflects the structure of the real-world task.

\subsection{Critical synthesis}

Across these strands of literature, several conclusions emerge consistently. First, validation protocol choice matters substantially, particularly when data are limited. Second, uncertainty in performance estimation is both methodologically important and difficult to quantify reliably. Third, model selection can itself distort evaluation, especially when noisy validation results are repeatedly exploited. Fourth, metric choice is inseparable from evaluation design in imbalanced settings. Finally, applied studies confirm that these are not merely theoretical concerns but real methodological risks in practical machine learning.

At the same time, the literature reveals an unresolved tension. Many papers successfully identify sources of instability, bias, or misinterpretation, but fewer integrate these insights into a single practical framework for evaluation and reporting. In particular, the connection between uncertainty-aware resampling, imbalance-sensitive metric selection, and transparent model comparison remains less clearly structured than the individual components themselves. This unresolved issue motivates the research gap developed in the next section.

\section{Research Gap and Objectives}

The literature reviewed above establishes that model evaluation under limited-data conditions is inherently fragile. Foundational studies have shown that cross-validation procedures differ in important statistical ways, that uncertainty in validation estimates is difficult to quantify reliably, and that model selection itself can introduce bias into reported performance. Additional work on imbalanced classification has shown that metric choice can strongly affect interpretation, while applied studies demonstrate that different validation designs can materially change the apparent stability of model performance.

However, despite these important contributions, several limitations remain. First, much of the existing literature focuses on identifying methodological problems rather than translating them into a coherent evaluation strategy for practical use. Theoretical work explains why uncertainty estimation is difficult, but often stops short of proposing a clearly structured framework for uncertainty-aware reporting in constrained-data settings. Second, the interaction between limited sample size and class imbalance is acknowledged in parts of the literature, but it is not always incorporated into a unified account of evaluation design. As a result, resampling strategy and metric choice are often discussed in parallel rather than as interconnected components of trustworthy model assessment.

A further limitation is that many studies still emphasise mean performance estimates more strongly than the broader question of interpretive reliability. In practical machine learning workflows, the central issue is rarely just whether one protocol yields a slightly higher average score than another. The more important question is whether the evaluation process provides a sufficiently stable, transparent, and defensible basis for deciding which model should be trusted. This is especially relevant when limited data magnifies the impact of sample composition, fold assignment, and minority-class representation.

Therefore, the main research gap identified in this portfolio is the lack of a clear and practically structured evaluation framework that brings together validation-protocol comparison, uncertainty-aware interpretation, and imbalance-sensitive metric selection for classical machine learning models under limited and imbalanced data conditions. Existing research provides strong insights into individual aspects of the problem, but less guidance on how these aspects should be integrated into a coherent methodological design.

This gap leads to the following research question:

\textbf{How reliable are common evaluation protocols for classical machine learning models under limited and imbalanced data, and can uncertainty-aware resampling strategies provide more trustworthy model assessment?}

To address this question, the portfolio is guided by four objectives:
\begin{enumerate}[label=\arabic*.]
    \item To review how current literature evaluates classical machine learning models in limited-data settings.
    \item To compare common evaluation protocols, including hold-out split, stratified k-fold cross-validation, repeated cross-validation, and bootstrap-based estimation.
    \item To assess how limited sample size and class imbalance affect the stability and trustworthiness of reported performance metrics.
    \item To propose a more transparent and uncertainty-aware evaluation framework for constrained-data settings.
\end{enumerate}

These objectives are intentionally methodological rather than algorithmic. The purpose of the study is not to maximise benchmark performance or introduce a novel predictive model. Instead, the goal is to improve the reliability and interpretability of model assessment under constrained-data conditions. In this sense, the expected contribution of the portfolio lies in evaluation design and reporting practice rather than in model innovation.

\section{Proposed Methodology}

The proposed study is designed to investigate how common evaluation protocols behave when classical machine learning models are assessed under limited and imbalanced data conditions. The methodological emphasis is deliberate. Rather than asking which algorithm achieves the highest score on a particular benchmark, the study asks how different validation procedures influence the stability, interpretation, and trustworthiness of performance estimates. This aligns directly with the research gap identified in the literature, namely the need for a more integrated framework combining validation-protocol comparison, uncertainty-aware interpretation, and imbalance-sensitive metric selection.

\subsection{Study design}

The study will adopt a comparative methodological design. A small set of classical machine learning models will be evaluated under multiple validation protocols, and the resulting performance estimates will be compared in terms of stability, sensitivity to sample composition, and implications for model ranking. The focus is therefore not on maximising predictive performance, but on examining the behaviour of evaluation procedures themselves.

This design is appropriate because it allows the study to isolate the role of evaluation strategy while avoiding unnecessary complexity from model architecture. Since the portfolio is concerned with reliability of assessment rather than novelty of modelling technique, a compact and interpretable design is preferable.

\subsection{Data strategy}

The proposed study will use classification datasets that reflect constrained-data conditions. These may include one or two publicly available real datasets with relatively small sample sizes, as well as down-sampled versions of larger datasets if needed to examine how stability changes as sample size decreases. Where possible, datasets with some degree of class imbalance will be prioritised, since imbalance is a common feature of limited-data applications and is central to the methodological focus of the project.

The purpose of this data strategy is not to claim domain-wide generalisability from a single dataset, but to create a setting in which the methodological effects of validation design can be observed clearly. A controlled comparison across small and potentially imbalanced datasets would allow the study to examine how strongly performance estimates depend on resampling choice and metric selection.

\subsection{Model selection}

The study will focus on a limited set of widely used classical machine learning models, such as Logistic Regression, Random Forest, and either Support Vector Machine or XGBoost. These models are suitable for several reasons. First, they are common reference points in applied machine learning and therefore make the findings easier to interpret. Second, they provide variation in modelling behaviour without introducing the additional complexity associated with deep learning architectures. Third, they allow the study to keep the methodological focus on evaluation procedures rather than on extensive hyperparameter engineering.

The intention is not to determine a universally best classifier, but to assess whether conclusions about model superiority remain stable across different validation designs.

\subsection{Evaluation protocols}

Four evaluation protocols will be compared:
\begin{itemize}
    \item a single hold-out split,
    \item stratified k-fold cross-validation,
    \item repeated stratified k-fold cross-validation,
    \item bootstrap-based performance estimation.
\end{itemize}

These protocols were selected because they represent a progression from simple but potentially unstable methods to more resampling-intensive approaches that may provide better insight into uncertainty. A single hold-out split is included as a baseline because it remains common in practice despite known instability in small datasets. Stratified k-fold cross-validation is included because it is a widely accepted standard and partially addresses class-composition issues. Repeated stratified k-fold extends this logic by examining whether repeated resampling improves stability and interpretive confidence. Bootstrap-based estimation is included because it offers an alternative resampling framework that may produce more informative uncertainty summaries in some settings.

Comparing these protocols will allow the study to assess not only differences in average reported performance, but also whether protocol choice changes the apparent stability of the estimate and the practical ranking of competing models.

\subsection{Performance metrics}

The study will report metrics that are appropriate for imbalanced classification and transparent interpretation. These are expected to include F1 score, Balanced Accuracy, AUROC, and AUPRC. In addition, uncertainty-related summaries such as mean performance across resamples, standard deviation, and where appropriate interval-style summaries will be considered. The inclusion of both central tendency and uncertainty measures is essential because the project is concerned not only with how well models perform, but also with how confidently those performance claims can be made.

Metric choice is particularly important under class imbalance. A model may achieve a strong overall ranking metric while performing poorly on the minority class, which may be the most practically important set of cases. For this reason, the proposed methodology treats metric selection as part of evaluation design rather than as a separate reporting detail.

\subsection{Analysis plan}

The analysis will focus on five main questions.

First, how stable are performance estimates under different evaluation protocols? This addresses the central concern that different validation strategies may produce substantially different levels of variability.

Second, how does limited sample size affect evaluation reliability? By using small or down-sampled datasets, the study will examine whether instability increases as the sample becomes more constrained.

Third, how does class imbalance affect metric interpretation? This part of the analysis will consider whether conclusions differ depending on whether performance is assessed through more general ranking measures or more imbalance-sensitive metrics.

Fourth, how consistent is model ranking across protocols? This question is important because in practice the goal is often to select one model rather than merely to report several scores. If ranking changes substantially across evaluation designs, this would indicate that the apparent superiority of a model is not robust.

Fifth, does uncertainty-aware reporting provide a more defensible basis for interpretation than single point estimates alone? This question connects the empirical comparison back to the broader methodological goal of transparency in constrained-data evaluation.

\subsection{Expected contribution}

The expected contribution of the proposed study is methodological. It is intended to provide a more structured basis for evaluating classical machine learning models under limited and imbalanced data conditions. Rather than proposing a new algorithm, the study aims to clarify how evaluation protocols, metric choice, and uncertainty summaries should be interpreted together when evidence is limited. The intended outcome is a practical recommendation-oriented framework for more trustworthy model assessment.

\subsection{Practical and ethical considerations within the methodology}

The methodological design also reflects practical and ethical concerns. Overconfident evaluation in small or imbalanced datasets can support claims that are not justified by the available evidence. This is particularly problematic in applications where model outputs may influence consequential decisions. For this reason, the proposed methodology emphasises caution in interpretation, explicit attention to uncertainty, and metric choices that reflect the actual prediction problem rather than reporting convenience.

Overall, the proposed methodology is designed to respond directly to the weaknesses identified in the literature: excessive reliance on point estimates, insufficient integration of imbalance into evaluation logic, and limited guidance on how to translate methodological warnings into a coherent assessment framework.

\section{Data, Ethics, and Practical Considerations}

The proposed study requires datasets that reflect constrained-data conditions and, where possible, some degree of class imbalance. The purpose of dataset selection is not to maximise benchmark performance, but to create a setting in which the methodological effects of validation design can be examined clearly. For this reason, relatively small classification datasets, or down-sampled versions of larger datasets, would be appropriate for the study. These data conditions are essential because the portfolio is concerned with how evaluation behaves when evidence is limited rather than when data abundance masks methodological weaknesses.

From a practical perspective, several considerations shape the feasibility of the proposed design. First, the datasets must support repeated resampling without becoming too small to permit meaningful comparison. Second, model complexity must remain proportionate to sample size, otherwise instability may be driven mainly by severe overfitting rather than by evaluation design itself. Third, metric choice must reflect the structure of the data, particularly where class imbalance makes overall performance summaries potentially misleading. These considerations reinforce the need for a methodology that is transparent not only about numerical results, but also about the limits of those results.

There are also important ethical implications. Unreliable evaluation can produce overconfident claims about model quality, which may in turn encourage inappropriate trust in a system that has not been assessed robustly. This concern is particularly significant in applied domains where model outputs may influence decisions affecting individuals, resources, or risk management. In such contexts, reporting only a single favourable performance estimate may conceal instability, weak minority-class performance, or sensitivity to arbitrary split decisions.

To address these concerns, the proposed framework emphasises three principles. First, uncertainty should be made visible rather than hidden behind point estimates alone. Second, metric choice should reflect the real prediction problem, particularly when minority-class performance is practically important. Third, conclusions should remain proportionate to the strength of the evidence. The proposed study therefore adopts a cautious reporting stance: the goal is not to maximise apparent model quality, but to provide a more defensible basis for interpretation under constrained-data conditions.

\section{Project Planning}

The development of this portfolio follows a structured two-week plan designed to connect topic scoping, literature review, methodological design, and final writing into a coherent workflow. The project begins with repository setup, research question refinement, and literature collection. This is followed by thematic clustering of papers, close reading of key methodological studies, and formal identification of the research gap. The second phase focuses on designing the proposed methodology, drafting the report sections, revising the argument, and preparing submission materials.

This sequencing is important because each stage depends on the previous one. The literature review informs the gap analysis; the gap analysis shapes the methodology; and the methodology must then be aligned with practical, ethical, and reporting considerations. A structured plan is therefore necessary not only for time management but also for intellectual coherence. Without such sequencing, the report would risk becoming a collection of loosely related observations rather than a well-argued portfolio.

The use of GitHub further supports this planning process by providing a transparent record of development. Planning notes, literature summaries, methodology sketches, and report drafts are stored as part of the repository so that the progression from early scoping to final written output is visible. This makes the repository part of the academic process rather than merely a storage location for the final PDF.

\section{Non-Technical Summary}

This project addresses a simple but important question: how can we tell whether a machine learning model is genuinely reliable when we do not have much data?

In many real-world settings, machine learning models are tested on relatively small datasets. The resulting scores can look precise and convincing, but they may depend heavily on how the data was split, which validation procedure was chosen, and which performance metric was reported. The problem becomes even more serious when one class is much rarer than another, because some commonly reported metrics may make a model look stronger than it really is.

The purpose of this project is therefore not to build a new model. Instead, it is to improve the way models are evaluated. The proposed work compares common validation procedures and considers how uncertainty should be interpreted and reported more carefully. In practical terms, the project aims to support evaluation approaches that are more stable, more transparent, and less likely to encourage unjustified confidence in model performance.

This matters because poor evaluation can lead to poor decisions. If a model appears strong only because of unstable testing or selective reporting, then claims about its usefulness may not be trustworthy. A better evaluation framework can therefore improve not only technical practice but also the quality of decisions made on the basis of machine learning systems.

\section{Conclusion}

This portfolio has examined the methodological challenge of evaluating classical machine learning models under limited and imbalanced data conditions. The literature shows that validation protocol choice, uncertainty estimation, model selection bias, and metric selection all play important roles in shaping how model performance is interpreted. It also shows that seemingly strong results may be unstable, overly optimistic, or insufficiently informative when evidence is limited.

In response to these issues, the report has identified a research gap in the lack of a clearly structured framework that integrates uncertainty-aware resampling, imbalance-sensitive metric choice, and transparent model comparison. The proposed study addresses this gap through a methodological design that compares common validation protocols and places greater emphasis on interpretive reliability rather than score maximisation alone.

The intended contribution of the portfolio is therefore not algorithmic innovation, but methodological clarification. By focusing on how models should be evaluated, rather than only on how they should be built, the project aims to support more trustworthy and more defensible model assessment in constrained-data settings.

\bibliographystyle{plain}
\bibliography{refs}

\end{document}
